Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves

### 2025-08-19 ms/ms.qmd
migrating main text to word

```{=latex}
\begin{center}
$^1$School of Life Sciences, University of Hawaiʻi, Mānoa, Hawaiʻi, USA\\
$^2$Department of Botany, University of Wisconsin, Madison, Wisconsin, USA\\
\medskip
*Corresponding author: cdmuir@wisc.edu\\
\medskip
ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
\end{center}
```

@@

The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts. Paradoxically, despite this photosynthetic advantage, most leaves are hypostomatous (stomata on lower surface only). Based on the observation that amphistomatous leaves are most common among herbaceous plants growing in sunny habitats, including many crop species, it has long been hypothesized that traits affecting CO$_2$ supply or demand amplify the advantage of amphistomy under high light intensity. We used a recently developed method to quantify amphistomy advantage, the photosynthetic rate of an amphistomatous leaf relative to an otherwise identical hypostomatous leaf, in `r n_acc` diverse populations representing `r n_sp` species of wild tomatoes (*Solanum* sect. *Lycopersicon* and sect. *Lycopersicoides*). Plants grown under high light intensity benefit more from amphistomy in terms of CO$_2$ assimilated for a given stomatal conductance than plants grown under low light, regardless of measurement light intensity. Furthermore, populations native to open habitats tend to benefit more from amphistomy compared to those from more closed habitats. Developmental plasticity and adaptation of leaves to high light intensity likely play a combined role in explaining the adaptive significance and distribution of amphistomy in nature. We estimate that hypostomatous leaves would expend 10-65\% more water to achieve the same photosynthetic rate as a comparable amphistomatous leaf, implying large fitness costs of amphistomy in ecological contexts where this trait is rare.

@@

<!-- Paragraph 1 -->
Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells. They are essential for balancing carbon gained per unit water lost and were an essential innovation in vascular plants, permitting them to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mcadam_stomata_2021; @clark_origin_2022]. Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@hetherington_role_2003; @de_boer_optimal_2016; @harrison_influence_2020]. Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for studying adaptation [@buckley_modelling_2013; @raven_selection_2002; @haworth_co-ordination_2013], inferring paleoclimate from fossil cuticles [@franks_new_2014; @mcelwain_paleoecology_2017; @the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023], predicting responses to climate change [@woodward_stomatal_1987; @liang_stomatal_2023; @chua_stomatal_2024; @lang_century-long_2024], and improving crops [@hofmann_impact_2025]. Stomatal function contributes to global carbon and water cycles [@berry_stomata:_2010] and therefore predicting future climate [@franks_stomatal_2017].

@@

*Acclimation hypothesis*: Photosynthetic induction to high light intensity typically involves increases in total leaf stomatal conductance (increased CO$_2$ supply), the concentration of active Rubisco, and electron transport capacity (increased CO$_2$ demand). If the acclimation hypothesis is correct, we predict that $\aahigh{} > \aalow{}$ for all populations regardless of native habitat or growth environment.

*Plasticity hypothesis*: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions [@poorter_metaanalysis_2019]. Plastic responses are likely adaptations to optimize photosynthesis at different light intensities in variable environments [@givnish_common-garden_2014]. Plastic changes in leaf anatomy and biochemistry could modulate \aax{} as a byproduct. Thicker or less porous leaves, both of which are associated with high leaf mass per area (LMA), will have lower \gias; leaves with increased total stomatal density and photosynthetic capacity have greater potential CO$_2$ supply and demand. Under the plasticity hypothesis, we predict that $\aasun{} > \aashade{}$ for all populations and light intensities.

@@

```{r, hypotheses, echo = FALSE}
#| label: fig-hypotheses
#| fig-align: center
#| fig-cap: "\\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.} The acclimatopn hypothesis predicts that \\aax{} is greater under high measurement light intensity (\\ppfdqty{2000}) than low measurement light intensity (\\ppfdequals{150}), regardless of growth light intensity or native plant area index (PAI). The plasticity hypothesis predicts that \\aax{} is greater for plants grown in sun than shade, regardless of measurement light intensity or native PAI. The constitutive hypothesis predicts that \\aax{} is greater for plants adapted to sunny habitats (lower native PAI) than shaded habitats, regardless of measurement light intensity or growth light intensity."

@@

<!-- Paragraph 5 (summarize methods) -->
We tested these hypotheses by comparing \aax{} among amphistomatous wild tomato species (@tbl-populations; @fig-sr-rxnorm; [@peralta_taxonomy_2008]) from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensities (low and high). We measured \aax{} on `r n_plant` individual plants from `r n_acc` populations (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024]. With this method, we directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at identical whole-leaf stomatal conductance (\gsw), we measure $A$ over a range of \gsw, inducing stomatal opening and closure by modulating humidity (see Materials and Methods for further details). We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024 and here (Materials and Methods). The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded data set derived from the Global Ecosystem Dynamics Investigation [GEDI; @burns_multi-resolution_2024] and georeferenced accession collection information from the Tomato Genetics Resource Center (@tbl-populations). The growth light intensities were \ppfdequals{`r ppfd_sun`} (sun treatment) and \ppfdqty{`r ppfd_shade`} (shade treatment) while all other environment conditions were nearly identical (see Materials and Methods). The high and low measurement light intensities were \ppfdequals{2000} (`r p_red_high`:`r p_blue_high` red:blue) and \ppfdequals{150} (`r p_red_low`:`r p_blue_low` red:blue), respectively.

@@

<!-- Paragraph 7 (summarize secondary results) -->
The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, contradicts the acclimation hypothesis, and provides modest support for the constitutive hypothesis. Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), may mediate the effect of growth light intensity on \aax{}. LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], qualitatively similar to plastic responses in many species [@poorter_metaanalysis_2019]. While LMA is weakly, albeit significantly (@tbl-fit_aa2), associated with individual-level \aax{} ([Fig. @fig-lma_aa]), the effect of growth light intensity on \aax{} is still predictive based on model comparison (@tbl-aa_loo1). However, the direct effect of the sun treatment on \aax{} was weaker when LMA was included in the model (@tbl-fit_aa1 vs. @tbl-fit_aa2), suggestive of a mediating role. Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed to  identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}. The fact that the \aax{} of sun plants was greater under low measurement light intensity supports a long-standing hypothesis that resistance to CO$_2$ diffusion is greater in the upper than lower portions of the leaf interior [@jones_effects_1972]. At low light intensity, photosynthesis is weighted toward the upper palisade where most light is intercepted. If resistance to CO$_2$ diffusion is high in the upper palisade, amphistomy may, unexpectedly, be particularly beneficial for sun leaves experiencing intermittent shade or cloud cover. Our study is limited in testing this because we could not directly measure the stomatal conductance ratio and intercellular resistance on each surface. Future experiments measuring \aax{} with a dual sided chamber [@wall_stomata_2022; @marquez_assessing_2023] can overcome these limitations.

```{r, lma_aa, echo = FALSE,warning=FALSE, message=FALSE} 
#| label: fig-lma_aa
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "\\textbf{Plastic changes in leaf mass per area (LMA) mediate the effect of sun- to shade-grown plants on amphistomy advantage (\\aax{}).} Each point is the estimated population-level LMA ($x$-axis) and \\aax{} ($y$-axis) for each population grown under high light intensity (\\ppfdequals{2000}) and low light intensity (\\ppfdequals{150}). The points are colored by growth light treatment. The solid line is the posterior median of the linear regression of individual-level LMA on $\\aahigh$  with the shaded region showing the 95% confidence ribbon; the dashed line is the same for $\\aalow$. Confidence intervals for each population estimate were omitted for visual clarity."

>>>>

```{=latex}
\begin{center}
$^1$School of Life Sciences, University of Hawaiʻi; Mānoa, Hawaiʻi, USA\\
$^2$Department of Botany, University of Wisconsin, Madison; Wisconsin, USA\\
\medskip
*Corresponding author: cdmuir@wisc.edu\\
\medskip
ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
\end{center}
```

@@

The presence of stomata on both leaf surfaces (amphistomy) increases photosynthesis by reducing the distance for CO$_2$ diffusion between stomata and chloroplasts. Paradoxically, most leaves are hypostomatous (stomata on lower surface), despite the photosynthetic advantage of amphistomy. Across `r n_acc` diverse populations of "wild tomatoes", leaves developed under high light intensity benefit more from amphistomy in terms of CO$_2$ assimilated for a given stomatal conductance than plants developed under low light. Furthermore, populations native to open habitats benefit more from amphistomy than those from more closed habitats. Thus, plasticity and adaptation together may explain why amphistomatous leaves are prevalent in sunny, open habitats, including many crops. Contrary to common assumptions, amphistomy can save water because hypostomatous leaves evaporate 10-65\% more to achieve the same photosynthetic rate.

 <!-- LONG FORM - save in case of resubmission -->
<!-- The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts. Paradoxically, despite this photosynthetic advantage, most leaves are hypostomatous (stomata on lower surface only). Based on the observation that amphistomatous leaves are most common among herbaceous plants growing in sunny habitats, including many crop species, it has long been hypothesized that traits affecting CO$_2$ supply or demand amplify the advantage of amphistomy under high light intensity. We used a recently developed method to quantify amphistomy advantage, the photosynthetic rate of an amphistomatous leaf relative to an otherwise identical hypostomatous leaf, in `r n_acc` diverse populations representing `r n_sp` species of wild tomatoes (*Solanum* sect. *Lycopersicon* and sect. *Lycopersicoides*). Plants grown under high light intensity benefit more from amphistomy in terms of CO$_2$ assimilated for a given stomatal conductance than plants grown under low light, regardless of measurement light intensity. Furthermore, populations native to open habitats tend to benefit more from amphistomy compared to those from more closed habitats. Developmental plasticity and adaptation of leaves to high light intensity likely play a combined role in explaining the adaptive significance and distribution of amphistomy in nature. We estimate that hypostomatous leaves would expend 10-65\% more water to achieve the same photosynthetic rate as a comparable amphistomatous leaf, implying large fitness costs of amphistomy in ecological contexts where this trait is rare. -->

@@

<!-- Paragraph 1 -->
Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells. They are essential for balancing carbon gained per unit water lost and were an essential innovation in vascular plants, permitting them to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mcadam_stomata_2021; @clark_origin_2022]. Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@hetherington_role_2003; @de_boer_optimal_2016; @harrison_influence_2020]. Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for studying adaptation [@buckley_modelling_2013; @raven_selection_2002; @haworth_co-ordination_2013], inferring paleoclimate from fossil cuticles [@franks_new_2014; @mcelwain_paleoecology_2017; @the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023], predicting responses to climate change [@woodward_stomatal_1987; @liang_stomatal_2023; @chua_stomatal_2024; @lang_century-long_2024], and improving crops [@hofmann_impact_2025]. Stomatal function contributes to global carbon and water cycles [@berry_stomata:_2010] and therefore to predicting future climate [@franks_stomatal_2017].

@@

*Acclimation hypothesis*: Photosynthetic induction to high light intensity typically involves increases in total leaf stomatal conductance (increased CO$_2$ supply), the concentration of active Rubisco, and electron transport capacity (increased CO$_2$ demand). If the acclimation hypothesis is correct, we predict that $\aahigh{} > \aalow{}$ for all populations regardless of native habitat or growth environment. $\aahigh{}$ is the \aax{} measured under high light intensity (\ppfdequals{2000}); $\aalow{}$ is the \aax{} measured under low light intensity (\ppfdequals{150}).

*Plasticity hypothesis*: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions [@poorter_metaanalysis_2019]. Plastic responses are likely adaptations to optimize photosynthesis at different light intensities in variable environments [@givnish_common-garden_2014]. Plastic changes in leaf anatomy and biochemistry could modulate \aax{} as a byproduct. Thicker or less porous leaves, both of which are associated with high leaf mass per area (LMA), will have lower \gias; leaves with increased total stomatal density and photosynthetic capacity have greater potential CO$_2$ supply and demand. Under the plasticity hypothesis, we predict that $\aasun{} > \aashade{}$ for all populations and light intensities. $\aasun{}$ is the \aax{} measured on sun leaves; $\aashade{}$ is the \aax{} measured on shade leaves.

@@

```{r, hypotheses, echo = FALSE}
#| label: fig-hypotheses
#| fig-align: center
#| fig-cap: "\\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.} The acclimation hypothesis predicts that \\aax{} is greater under high measurement light intensity (\\ppfdequals{2000}) than low measurement light intensity (\\ppfdequals{150}), regardless of growth light intensity or native plant area index (PAI). The plasticity hypothesis predicts that \\aax{} is greater for plants grown in sun than shade, regardless of measurement light intensity or native PAI. The constitutive hypothesis predicts that \\aax{} is greater for plants adapted to sunny habitats (lower native PAI) than shaded habitats, regardless of measurement light intensity or growth light intensity."

@@

<!-- Paragraph 5 (summarize methods) -->
We tested these hypotheses by comparing \aax{} among amphistomatous wild tomato species (@tbl-populations; @fig-sr-rxnorm; [@peralta_taxonomy_2008]) from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensities (low and high). We measured \aax{} on `r n_plant` individual plants from `r n_acc` populations (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024]. With this method, we directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at identical whole-leaf stomatal conductance (\gsw), we measure $A$ over a range of \gsw, inducing stomatal opening and closure by modulating humidity (see supplementary text for further details). We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024 and here (Supplementary Materials). The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded data set derived from the Global Ecosystem Dynamics Investigation [GEDI; @burns_multi-resolution_2024] and georeferenced accession collection information from the Tomato Genetics Resource Center (@tbl-populations). The growth light intensities were \ppfdequals{`r ppfd_sun`} (sun treatment) and \ppfdqty{`r ppfd_shade`} (shade treatment) while all other environment conditions were nearly identical (see Materials and Methods). The high and low measurement light intensities were \ppfdequals{2000} (`r p_red_high`:`r p_blue_high` red:blue) and \ppfdequals{150} (`r p_red_low`:`r p_blue_low` red:blue), respectively.

@@

<!-- Paragraph 7 (summarize secondary results) -->
The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, contradicts the acclimation hypothesis, and provides modest support for the constitutive hypothesis. Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), may mediate the effect of growth light intensity on \aax{}. LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], qualitatively similar to plastic responses in many species [@poorter_metaanalysis_2019]. While LMA is weakly, albeit significantly (@tbl-fit_aa2), associated with individual-level \aax{} ([Fig. @fig-lma_aa]), the effect of growth light intensity on \aax{} is still predictive based on model comparison using information criteria (@tbl-aa_loo1). However, the direct effect of the sun treatment on \aax{} was weaker when LMA was included in the model (@tbl-fit_aa1 vs. @tbl-fit_aa2), suggestive of a mediating role. Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed to  identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}. The fact that the \aax{} of sun plants was greater under low measurement light intensity supports a long-standing hypothesis that resistance to CO$_2$ diffusion is greater in the upper than lower portions of the leaf interior [@jones_effects_1972]. At low light intensity, photosynthesis is weighted toward the upper palisade where most light is intercepted. If resistance to CO$_2$ diffusion is high in the upper palisade, amphistomy may, unexpectedly, be particularly beneficial for sun leaves experiencing intermittent shade or cloud cover. Our study is limited in testing this because we could not directly measure the stomatal conductance ratio and intercellular resistance on each surface. Future experiments measuring \aax{} with a dual sided chamber [@wall_stomata_2022; @marquez_assessing_2023] can overcome these limitations.

```{r, lma_aa, echo = FALSE,warning=FALSE, message=FALSE} 
#| label: fig-lma_aa
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "\\textbf{Plastic changes in leaf mass per area (LMA) partially mediate the effect of sun- to shade-grown plants on amphistomy advantage (\\aax{}).} Each point is the estimated LMA ($x$-axis) and \\aax{} ($y$-axis) for each population grown under high light intensity (\\ppfdequals{2000}) and low light intensity (\\ppfdequals{150}). The points are colored by growth light treatment. The solid line is the posterior median of the linear regression of individual-level LMA on $\\aahigh$  with the shaded region showing the 95% confidence ribbon; the dashed line is the same for $\\aalow$. Confidence intervals for each population estimate were omitted for visual clarity."
